---
title: "Boqian_Mao_Biostat_620_Hw4"
author: "Boqian Mao"
format: html
editor: visual
embed-resources: true
fig-width: 5
fig-height: 5
theme: cosmo
---

## Q1

```{r}
# Use the functions with this data
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)
```

```{r}
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i, j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}

fun2alt <- function(mat) {
  t(apply(mat, 1, cumsum))
}

out_original <- fun2(dat)
out_fast <- fun2alt(dat)

all.equal(out_original, out_fast)
```

```{r}
# install.packages("microbenchmark")
library(microbenchmark)

mb_q1 <- microbenchmark(
  original = fun2(dat),
  faster   = fun2alt(dat),
  times = 100L
)

mb_q1
```
```{r}
boxplot(mb_q1, main = "Q1 Runtime Comparison")
```

## Q2
```{r}
fun3 <- function(x) {
  out <- numeric(length(x))
  cumsum_x <- 0
  for (i in seq_along(x)) {
    if (x[i] == 0) {
      cumsum_x <- 0
    } else {
      cumsum_x <- cumsum_x + x[i]
    }
    out[i] <- cumsum_x
  }
  out
}

fun3alt <- function(x) {
  cs <- cumsum(x)
  reset <- cummax((x == 0) * cs)
  cs - reset
}

set.seed(2315)
x <- sample(c(0, -5:5), size = 100000, replace = TRUE)


library(microbenchmark)
mb_q2 <- microbenchmark(
  original = fun3(x),
  faster   = fun3alt(x),
  times = 100L
)

print(mb_q2)

# visualization
boxplot(mb_q2, main = "Q2 Runtime Comparison")
```

## Q3
```{r}
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Here is an example of the run
set.seed(156)
sim_pi(1000) # 3.132
```

```{r}
set.seed(1231)

time_serial <- system.time({
  ans_serial <- unlist(lapply(1:4000, sim_pi, n = 10000))
  serial_mean <- mean(ans_serial)
  print(serial_mean)
})
```

```{r}
library(parallel)

n_cores <- max(1, detectCores() - 1)
cl <- makeCluster(n_cores)

clusterExport(cl, "sim_pi")
clusterSetRNGStream(cl, 1231)

time_parallel <- system.time({
  ans_parallel <- unlist(parLapply(cl, 1:4000, sim_pi, n = 10000))
  parallel_mean <- mean(ans_parallel)
  print(parallel_mean)
})

print(time_parallel)

stopCluster(cl)
```

```{r}
library(ggplot2)

df_pi <- data.frame(
  estimate = c(ans_serial, ans_parallel),
  method = rep(c("Serial", "Parallel"),
               times = c(length(ans_serial), length(ans_parallel)))
)

ggplot(df_pi, aes(x = estimate, fill = method)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = pi, linetype = "dashed") +
  labs(
    title = "Distribution of Pi Estimates",
    x = "Estimated Pi",
    y = "Density"
  ) +
  theme_minimal()

df_time <- data.frame(
  method = c("Serial", "Parallel"),
  elapsed = c(time_serial["elapsed"], time_parallel["elapsed"])
)

ggplot(df_time, aes(x = method, y = elapsed, fill = method)) +
  geom_col() +
  labs(
    title = "Runtime Comparison",
    x = "Method",
    y = "Elapsed Time (seconds)"
  ) +
  theme_minimal()
```
These plots show that the distributions of the estimated values of pi for both methods are similar and they have very similar ranges and centers, indicating that the parallelization changes speed but not the overall simulation results.

## SQL
```{r}
library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(RSQLite::SQLite(), ":memory:")

# Read in tables
surveys <- read.csv("surveys.csv")
species <- read.csv("species.csv")
plots <- read.csv("plots.csv")

# Copy data.frames to database
dbWriteTable(con, "surveys", surveys, overwrite = TRUE)
dbWriteTable(con, "species", species, overwrite = TRUE)
dbWriteTable(con, "plots", plots, overwrite = TRUE)
```

## Q4
```{r}
dbGetQuery(con, "
  SELECT year,
         month,
         day,
         species_id,
         ROUND(weight / 1000.0, 2) AS weight_kg
  FROM surveys
  WHERE year = 1999
  ORDER BY species_id;
")
```

## Q5
```{r}
dbGetQuery(con, "
  SELECT SUM(weight) AS total_weight,
         AVG(weight) AS average_weight,
         MIN(weight) AS minimum_weight,
         MAX(weight) AS maximum_weight
  FROM surveys;
")
```
```{r}
dbGetQuery(con, "
  SELECT SUM(weight) AS total_weight,
         AVG(weight) AS average_weight,
         MIN(weight) AS minimum_weight,
         MAX(weight) AS maximum_weight
  FROM surveys
  WHERE weight BETWEEN 5 AND 10;
")
```
## Q6
```{r}
q6_result <- DBI::dbGetQuery(con, "
  SELECT year,
         species_id,
         COUNT(*) AS count_per_year_species,
         ROUND(AVG(weight), 2) AS avg_weight
  FROM surveys
  GROUP BY year, species_id
  ORDER BY year, species_id;
")

q6_result
```

## Q7
```{r}
dbGetQuery(con, "
  SELECT taxa,
         COUNT(*) AS num_species
  FROM species
  GROUP BY taxa
  HAVING COUNT(*) > 10
  ORDER BY num_species DESC;
")
```

## Q8
```{r}
q8_result <- DBI::dbGetQuery(con, "
  SELECT s.plot_id,
         sp.genus,
         COUNT(*) AS num_animals
  FROM surveys AS s
  INNER JOIN species AS sp
    ON s.species_id = sp.species_id
  GROUP BY s.plot_id, sp.genus
  ORDER BY s.plot_id ASC, num_animals DESC;
")

q8_result
```

## Q9
```{r}
dbGetQuery(con, "
  SELECT sp.species_id,
         sp.species,
         p.plot_type,
         COUNT(*) AS num_specimens
  FROM surveys AS s
  INNER JOIN species AS sp
    ON s.species_id = sp.species_id
  INNER JOIN plots AS p
    ON s.plot_id = p.plot_id
  WHERE sp.species_id != 'UN'
  GROUP BY sp.species_id, sp.genus, sp.species, p.plot_type
  ORDER BY sp.species_id, p.plot_type;
")
```

## Q10
```{r}
dbGetQuery(con, "
  SELECT sp.species_id,
         sp.genus,
         sp.species,
         ROUND(AVG(s.weight), 2) AS avg_weight
  FROM surveys AS s
  INNER JOIN species AS sp
    ON s.species_id = sp.species_id
  WHERE sp.taxa = 'Rodent'
  GROUP BY sp.species_id, sp.genus, sp.species
  ORDER BY avg_weight DESC;
")
```

## Q11
```{r}
library(ggplot2)
library(plotly)

q8_subset <- subset(q8_result, plot_id %in% 1:5)

p <- ggplot(q8_subset, aes(x = factor(plot_id), y = num_animals, fill = genus)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Number of Animals by Genus for Plot IDs 1–5",
    x = "Plot ID",
    y = "Number of Animals",
    fill = "Genus"
  ) +
  theme_minimal()

ggplotly(p)
```
## Q12
```{r}
library(plotly)
library(dplyr)

q6_clean <- q6_result %>%
  filter(!is.na(avg_weight)) %>%
  group_by(species_id) %>%
  filter(n() >= 5) %>%
  ungroup()

plot_ly(
  data = q6_clean,
  x = ~year,
  y = ~avg_weight,
  color = ~species_id,
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "Species: %{color}<br>",
    "Year: %{x}<br>",
    "Avg weight: %{y:.2f}<extra></extra>"
  )
) %>%
  layout(
    title = "Average Species Weight Over Time",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average Weight"),
    legend = list(title = list(text = "Species ID"))
  )
```
I used the q6_result table from Question 6 and removed rows with missing average weights before plotting. To reduce noise, I filtered the data to include only species with at least 5 yearly observations. This makes the interactive line chart easier to read while still showing temporal trends in average species weight.
