---
title: "Boqian_Mao_Biostat_620_Lab3"
author: "Boqian Mao"
format: html
editor: visual
embed-resources: true
fig-width: 5
fig-height: 5
theme: cosmo
---

## 1. Read in the data

```{r}
met <- read.csv("met_all.gz")
```

## 2. Check the dimensions, headers, footers

```{r}
dim(met)
```

2377343 rows and 30 columns

## 3. Take a look at all the variables (hint: use the str function).

```{r}
str(met)
```

## 4. Take a closer look at the key variables

```{r}
summary(met)
```

```{r}
met$elev[met$elev %in% c(9999)] <- NA
met$temp[met$temp %in% c(-40)] <- NA
met_new <- na.omit(met)
summary(met)
```

```{r}
max(met$elev, na.rm = TRUE)
min(met$temp, na.rm = TRUE)
```

At elevation 4113 is the highest weather station. The nwq lowest temperature is -17.2C.

## 5. Check the data against an external data source.

```{r}
met_clean <- subset(met, !is.na(temp))
sus <- subset(met_clean, temp == -17.2)
sus
```

It locates at Colorado near Colorado Springs/El Paso County.

## 6. Calculate summary statistics

```{r}
max_stat <- subset(met, elev == 4113)
vars <- met[, c("temp", "wind.sp", "hour", "day")]
cor(vars, use = "complete.obs")
```

## 7. Exploratory graphs

```{r}
par(mfrow = c(1, 3))

hist(met$elev,
     main = "Histogram of Elevation",
     xlab = "Elevation")

hist(met$temp,
     main = "Histogram of Temperature",
     xlab = "Temperature")

hist(met$wind.sp,
     main = "Histogram of Wind Speed",
     xlab = "Wind Speed")
```

The histogram of elevation shows a right skewed distribution. Temperature appears more symmetrically distributed, and wind speed is strongly right skewed. These distributions suggest that log-transformations of elevation and wind speed may be useful for later analyses.

```{r}
library(leaflet)

leaflet(max_stat) %>%
  addProviderTiles("OpenStreetMap") %>%
  addCircles(lat = ~lat,
             lng = ~lon,
             opacity = 1,
             fillOpacity = 1,
             radius = 10)
```

```{r}
library(lubridate)

max_stat$date <- with(max_stat,
                      ymd_h(paste(year, month, day, hour, sep = "-")))

summary(max_stat$date)

max_stat <- max_stat[order(max_stat$date), ]

head(max_stat)
```

```{r}
plot(max_stat$date,
     max_stat$temp,
     type = "l",
     xlab = "Date",
     ylab = "Temperature",
     main = "Temperature Time Series at Highest Elevation Station")
```

```{r}
plot(max_stat$date,
     max_stat$wind.sp,
     type = "l",
     xlab = "Date",
     ylab = "Wind Speed",
     main = "Wind Speed Time Series at Highest Elevation Station")
```

The temperature time series shows clear temporal variability, and values appear to vary smoothly over time. The wind speed time series exhibits greater short term variability, with frequent spikes and rapid changes. This pattern is consistent with wind being more volatile than temperature.

## 8. Ask questions

After exploring the dataset, several questions are about how the data were collected and structured. It is unclear whether weather stations are evenly distributed across the country or if certain areas are overrepresented, which could influence relationships between elevation, temperature, and wind speed. Also, the strong skewness in elevation and wind speed raises the question of whether extreme values from a small number of stations may disproportionately affect correlations and summary statistics.
