---
title: "Boqian_Mao_Biostat_620_Lab7"
author: "Boqian Mao"
format: html
editor: visual
embed-resources: true
fig-width: 5
fig-height: 5
theme: cosmo
---

## Question 1
```{r}
website <- xml2::read_html("https://pubmed.ncbi.nlm.nih.gov/?term=sars-cov-2")

counts <- xml2::xml_find_first(website, "//div[contains(@class,'results-amount')]/span[contains(@class,'value')]")

counts <- as.character(counts)

totalcount <- stringr::str_extract(counts, "[0-9][0-9.,]*")

totalcount <- gsub("[.,]", "", totalcount)

totalcount <- as.numeric(totalcount)
print(totalcount)
```

## Question 2
```{r}
# read in text, each line is a separate character
abstracts <- readLines('abstract-sars-cov-2-set.txt', warn = FALSE)
# combine all text into one character
abstracts <- paste(abstracts, collapse = "\n")
# split the text whenever 3 new lines occur in a row (indicating two blank lines)
abstracts <- strsplit(abstracts, "\n\n\n")
# replace any remaining "\n" symbols with spaces
abstracts <- gsub("\n", " ", abstracts)
# replace multiple spaces with single space
abstracts <- gsub(" +", " ", abstracts)
```

## Question 3
```{r}
library(stringr)

institution <- str_extract_all(
  abstracts,
  "\\b(?:University of [A-Za-z][A-Za-z .&'-]{2,}|[A-Za-z][A-Za-z .&'-]{2,} (?:State )?University|[A-Za-z][A-Za-z .&'-]{2,} Institute of [A-Za-z][A-Za-z .&'-]{2,})\\b"
)

institution <- unlist(institution)
table(institution)
```
The most frequent institution is Michigan State University (10), followed by University of Manitoba (3) and University of Michigan Medical School (3). To improve the results, I would normalize institution names to merge sub units into a parent university, and refine the regex to capture full institution names and avoid repeated matches.

## Question 4
```{r}
library(stringr)
abstracts <- readLines("abstract-sars-cov-2-set.txt", warn = FALSE)
abstracts <- paste(abstracts, collapse = "\n")
abstracts <- strsplit(abstracts, "\n\n\n")[[1]] 

line1   <- str_extract(abstracts, "(?m)^\\d+\\..*$")                 
journal <- str_replace(line1, "(?m)^\\d+\\.\\s*", "")              
journal <- str_replace(journal, "\\s20(20|21)\\b.*$", "")           

titles <- sapply(abstracts, function(x){
  unlist(strsplit(x, split = "\n\n"))[2]
}, USE.NAMES = FALSE)
titles <- gsub("\n", " ", titles)
titles <- gsub(" +", " ", titles)

authors <- sapply(abstracts, function(x){
  unlist(strsplit(x, split = "\n\n"))[3]
}, USE.NAMES = FALSE)
authors <- gsub("\n", " ", authors)
authors <- gsub(" +", " ", authors)


affiliations <- str_extract(abstracts, "(?s)\\n\\nAuthor information:\\n.*?\\n\\n")

affiliations <- str_replace(affiliations, "(?s)^\\n\\nAuthor information:\\n", "")
affiliations <- str_replace(affiliations, "(?s)\\n\\n$", "")
affiliations <- gsub("\n", " ", affiliations)
affiliations <- gsub(" +", " ", affiliations)

papers <- data.frame(
  journal = journal,
  title = titles,
  authors = substr(authors, 1, 100),
  affiliations = substr(affiliations, 1, 100),
  stringsAsFactors = FALSE
)

knitr::kable(papers[1:5, ])
```

